A thermodynamic view of dusty protoplanetary disks I: axisymmetric stability

### 2017-04-26 model.tex
new fig to illustrate phase lag

In general $\widetilde{\bm{v}}$ obeys a complicated evolutionary equation
\citep[see, e.g.][]{youdin05a}. However, for tightly-coupled dust
particles with $\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency, we can use the 
`terminal velocity approximation' to set %The dust-gas velocity difference 
%Then so that the dust-gas velocity
%difference is 
\begin{align}
  \widetilde{\bm{v}} = \frac{\nabla
    P}{\rhog}\tstop 
\end{align}
\citep{jacquet11}. This equation implies that dust particles
drift towards pressure maxima.

@@

We include the gravitational potential of the mixture 
$\psi$ for completeness but will neglect it throughout most of this
work ($\psi \equiv 0$) unless explicitly stated otherwise.

@@

Notice Eq. \ref{eos} resembles an ideal gas equation of state but with  
a reduced temperature $\tilde{T} = 
T_\mathrm{ref}\left(1-\tepsilon\right)$. %sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
This effective mixture  
temperature is lowered with dust-loading. 
%   \tepsilon\right)}$. 
%then the equation of state has the standard
%form $P=c_{s,\mathrm{eff}}^2\rho$. 
Since $\tepsilon$ typically decrease away from the midplane, we expect
vertically isothermal dusty disks to in fact behave as if it had a
temperture \emph{increasing} away from $z=0$.

@@

%In this form, the function $\mathcal{C}$ can be interpreted
%as a cooling term.    
If we denote 
\begin{align}
  \bm{F} \equiv  - \frac{\nabla P}{\rho}
\end{align}
for the pressure forces, then
\begin{align*}
  \mathcal{C} = - c_s^2\nabla \left( \tepsilon \tstop \rho \bm{F}
  \right), 
\end{align*}
which is in the same form as cooling by radiative diffusion. In protoplanetary
disks the heat flux, proportional to $\bm{F}$, is directly radially
outwards and vertically upwards. This is simply a reflection of
particle drift (inwards and downwards). Particle flux into a region contributes
to `cooling' of that region because the effective temperature is
lowered.

@@

The specific entropy of an ideal gas is 
$S = C_p\ln{\left(P^{1/\gamma}/\rhog\right)}$, where $C_p$ is the heat 
capacity at constant pressure.

@@

Positive work is done by a fluid parcel if $-\omega \imag\left(\Delta  
P\Delta\rho^*\right)>0$. Without loss of generality, take $\omega>0$
and consider a mass element with $\Delta\rho = 1$. Then positive work 
requires $\imag\left(\Delta
P\right)<0$.  This corresponds to pressure perturbations \emph{lagging}
behind that in density (or displacements).%, since the phase of
%the assumed perturbations increase in time.  
%$\imag\left(\Delta\rho\right)=0$.

>>>>

In general $\widetilde{\bm{v}}$ obeys a complicated evolutionary equation
\citep[see, e.g.][]{youdin05a}. However, for tightly-coupled dust
particles with $\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency, we can use the 
`terminal velocity approximation' to set %The dust-gas velocity difference 
%Then so that the dust-gas velocity
%difference is 
\begin{align}
  \widetilde{\bm{v}} = \frac{\nabla
    P}{\rhog}\tstop 
\end{align}
\citep{jacquet11}. This equation reflect the physical result that dust
particles tend to drift towards pressure maxima.

@@

The gravitational potential of the mixture 
$\psi$ is included here for completeness, but we will mostly neglect
it ($\psi \equiv 0$) unless explicitly stated otherwise.

@@

Notice Eq. \ref{eos} resembles an ideal gas equation of state but with  
a reduced temperature $\widetilde{T} = 
T_\mathrm{ref}\left(1-\tepsilon\right)$. %sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
This reduced temperature is lowered with dust-loading. 
%   \tepsilon\right)}$. 
%then the equation of state has the standard
%form $P=c_{s,\mathrm{eff}}^2\rho$. 
Since $\tepsilon$ typically decrease away from the midplane, we expect
vertically isothermal dusty disks to in fact behave as if it had a
temperture \emph{increasing} away from $z=0$.

@@

%In this form, the function $\mathcal{C}$ can be interpreted
%as a cooling term.    
If we denote 
\begin{align}
  \bm{F} \equiv  - \frac{\nabla P}{\rho}
\end{align}
for the pressure forces, then
\begin{align*}
  \mathcal{C} = - c_s^2\nabla \left( \tepsilon \tstop \rho \bm{F}
  \right), 
\end{align*}
which is in the same form as cooling by radiative diffusion. In protoplanetary
disks the corresponding heat flux, proportional to $\bm{F}$, is
directly radially outwards and vertically upwards. This is simply a
reflection of particle drift towards increasing pressure (inwards and 
downwards). Particle flux into a region contributes to `cooling' of
that region because the reduced temperature is lowered.

@@

The specific entropy of an ideal gas is 
$S = C_P\ln{\left(P^{1/\gamma}/\rhog\right)}$, where $C_P$ is the heat 
capacity at constant pressure.

@@

Positive work is done by a fluid parcel if $-\omega \imag\left(\Delta  
P\Delta\rho^*\right)>0$. Without loss of generality, take $\omega>0$
and consider a mass element with $\Delta\rho = 1$. Then positive work 
requires $\imag\left(\Delta
P\right)<0$.  This corresponds to pressure perturbations \emph{lagging}
behind that in density (or displacements), see Fig. \ref{lag_cartoon}
for the case of dusty gas.

%, since the phase of
%the assumed perturbations increase in time.  
%$\imag\left(\Delta\rho\right)=0$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 5cm 0cm 0cm]{figures/lag}
  \caption{Phase relation for overstable modes in isothermal dusty
    gas. Such modes require (Lagrangian) oscillations in the gas
    pressure, which is directly proportional to the gas density, to
    lag behind that in dust density. 
    \label{lag_cartoon}
  }
\end{figure}

### 2017-04-26 paper.tex
abstract slight edit
new fig to illustrate phase lag

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to entropy conservation
  following an adiabatic gas.  
  Motivated by this similarity, we develop a thermodynamical
  framework to study dusty gas dynamics in protoplanetary disks. We  
 show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to deduce that   
\begin{inparaenum}[1)] 
\item 
 the settling of perfectly coupled dust cannot cause axisymmetric 
 instabilties, however thin the dust layer becomes; 
\item 
  the streaming instability neccessarily involves a pressure response
  that lags dust density; 
\item dust-loading generally stabilizes the vertical shear
  instability, because dust-gas coupling introduces buoyancy. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss future applications of the dusty/adiabatic gas
correspondence in finding dusty analogs of other hydrodynamic
phenomena (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), and to simulate dusty protoplanetary disks
using purely hydrodynamic codes. %We briefly describe extending this
%equivalence principle to polytropic and adiabatic gas.

@@

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas friction introduces phenomena
that are absent in pure gas disks. Important examples include the
streaming instability 
\citep{youdin05a,youdin07b,johansen07}, secular gravitational
instabilities \citep[SGI,][]{ward00,youdin11,michikoshi12,takahashi14} and
Kelvin-Helmholtz instabilities
\citep{chiang08,barranco09,lee10}. These lead to structure and
planetesimal formation, as well as turbulence. 
More recently, new 
dusty instabilities have also been uncovered through numerical
simulations \citep{loren15,loren16,lamb16}, but these are not yet well
understood.

@@

By re-casting the dusty problem into identical form as standard  
hydrodynamics, we can apply well-known methods to generalize previous stability 
analyses for pure gas disks to the dusty case. %We explicitly demonstrate this for
%perfectly coupled dust in strictly and locally isothermal disks. 
We show that strictly isothermal protoplanetary disks with
perfectly-coupled dust are generally
stable to axisymmetric perturbations, regardless of how  thin the dust
layer is. We put forward a thermodynamical interpretation of
dust-gas drag instabilities, and apply it to the
streaming instability \citep[SI, ][]{youdin05a,youdin07b,jacquet11}. 
We also study the effect of dust-loading on the vertical shear
instability \citep[VSI,][]{nelson13,lin15,barker15}.

This paper is organized as follows. We first
explain in \S\ref{grow_osc} a general result of fluid dynamics that
underpins the rest of our study. We describe how 
a pressure-density phase lag in fluids can lead to 
instability. Thus finding an analogy between dusty gas and
pure hydrodynamics allow us to interpret dusty-gas dynamics
within a thermodynamics framework.

We begin our formal discussion in \S\ref{setup} by listing the 
basic equations describing tightly-coupled dusty gas. We show that for a
fixed equation of state, these are in 
fact the usual hydrodynamic equations in disguise. 
We describe equilibrium solutions in \S\ref{eqm} that are used in
explicit calculations. In \S\ref{limits} we discuss 
general stability properties of dusty disks. We then move onto
explicit stability analyses. In  
\S\ref{linear_problem} we list the linearized equations and show that 
they reproduce previous results on SGI and SI. In \S\ref{results} we
extend the VSI to dusty disks. We   
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

@@

From $A$ to $B$, a fluid parcel is expanding to return to 
equilbrium, but pressure (i.e. restoring) forces is still 
increasing. This over-compensation causes the parcel to expand
beyond the maximum volume of the previous cycle. 
Similarly, from $C$ to $D$ the parcel is already contracting towards
equilibrium, but pressure 
forces are still dropping. This allows the  
parcel's contraction to over-shoot the maximum density attained in the
previous cycle. 
The overall positive work done leads to growth in the oscillation amplitude. 
This effect is similar to being pushed downwards on a swing when 
descending.

@@

There are several situations where the pressure and density of a fluid
are not in phase. The obvious case is if the fluid is subject to
external heating/cooling. For example, in strongly irradiated
protoplanetary disks the
disk temperature  $T(r)$ is time-independent but varies with the cylindrical radius
$r$ from the star \citep{chiang97}. Since $P\propto \rhog T$ for an ideal gas, the pressure of a gas parcel varies with 
position as well. There is no reason to expect $P$ and $\rhog$ to
be in phase. In fact, we will show that this is a fundamental property of 
the vertical shear instability \citep{lin15}.

>>>>

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to entropy conservation
  following an adiabatic gas.  
  Motivated by this similarity, we develop a thermodynamical
  framework to study dusty gas dynamics in protoplanetary disks. We  
 show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to deduce that   
\begin{inparaenum}[1)] 
\item 
 perfectly coupled dust cannot cause axisymmetric 
 instabilties, however thin the dust layer is; 
\item 
  the streaming instability neccessarily involves a pressure response
  that lags dust density; 
\item dust-loading generally stabilizes the vertical shear
  instability, because dust-gas coupling introduces buoyancy. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss future applications of the dusty/adiabatic gas
correspondence in finding dusty analogs of other hydrodynamic
phenomena (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), and to simulate dusty protoplanetary disks
using purely hydrodynamic codes. %We briefly describe extending this
%equivalence principle to polytropic and adiabatic gas.

@@

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas friction introduces phenomena
that are absent in pure gas disks. Important examples include the
streaming instability 
\citep[SI,][]{youdin05a,youdin07b,johansen07}, secular gravitational
instabilities \citep[SGI,][]{ward00,youdin11,michikoshi12,takahashi14} and
Kelvin-Helmholtz instabilities
\citep{chiang08,barranco09,lee10}. These lead to structure and
planetesimal formation, as well as turbulence. 
More recently, new 
dusty instabilities have also been uncovered through numerical
simulations \citep{loren15,loren16,lamb16}, but these are not yet well
understood.

@@

By re-casting the dusty problem into identical form as standard  
hydrodynamics, we can apply well-known methods to generalize previous stability 
analyses for pure gas disks to the dusty case. %We explicitly demonstrate this for
%perfectly coupled dust in strictly and locally isothermal disks. 
We show that strictly isothermal protoplanetary disks with
perfectly-coupled dust are generally 
stable to axisymmetric perturbations, regardless of how  thin the dust
layer is. We put forward a thermodynamical interpretation of
dust-gas drag instabilities, and apply it to the streaming instability.   
We also study the effect of dust-loading on the vertical shear
instability \citep[VSI,][]{nelson13,lin15,barker15}.

This paper is organized as follows. We first
explain in \S\ref{grow_osc} a general result of fluid dynamics that
underpins the rest of our study. We describe how 
a pressure-density phase lag in fluids can lead to 
instability. Consequently, an analogy between dusty gas and  
pure hydrodynamics allow us to interpret dusty-gas instabilities 
within a thermodynamics framework.

We begin our formal discussion in \S\ref{setup} by listing the 
basic equations describing tightly-coupled dusty gas. We show that for a
fixed equation of state, these are in 
fact the usual hydrodynamic equations in disguise. 
We describe equilibrium solutions in \S\ref{eqm} that are used in
explicit calculations. In \S\ref{limits} we discuss 
general stability properties of dusty disks. We then move onto
explicit stability analyses. In  
\S\ref{linear_problem} we list the linearized equations and show that 
they reproduce previous results for SGI and SI. In \S\ref{results} we
extend the VSI to dusty disks. We   
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

@@

From $A$ to $B$, a fluid parcel is expanding to return to 
equilbrium, but pressure is still 
increasing. This over-compensation will cause the parcel to expand
beyond the maximum volume of the previous cycle. 
Similarly, from $C$ to $D$ the parcel is already contracting towards
equilibrium, but pressure 
is still dropping. This allows the  
parcel's contraction to over-shoot the maximum density attained in the
previous cycle. 
The overall positive work done leads to growth in the oscillation amplitude. 
This effect is similar to being pushed downwards on a swing when 
descending.

@@

There are several situations where the pressure and density of a fluid
are not in phase. The obvious case is if the fluid is subject to
external heating/cooling. For example, in strongly irradiated
protoplanetary disks the
disk temperature  $T(r)$ is time-independent but varies with the cylindrical radius
$r$ from the star \citep{chiang97}. Since $P\propto \rhog T$ for an
ideal gas, the pressure of a gas parcel varies with  
position as well. There is no reason to expect $P$ and $\rhog$ to
be in phase as a gas parcel oscillates between different radii. In
fact, we will show that this is a fundamental property of the vertical
shear instability \citep{lin15}.
